---
title: 'Reproducible Research: Peer Assessment 2'
author: "Victor Bernal"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: "html_document"
---

```{r setup, include=FALSE, echo=TRUE, tidy=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis

Severe weather events can cause public health (e.g. fatalities, injuries) and economic problems (e.g. property and crops damage). 
Thus the interest to identify the events with the largest human and economic impact. 
This Report uses the Storm Database of the U.S. National Oceanic and Atmospheric Administration (NOAA) 
The events in the database start in the year 1950 and end in November 2011. For more details <https://www.coursera.org/learn/reproducible-research/peer/OMZ37/course-project-2>.

# Loading and preprocessing the data
```{r, cache = TRUE}
fileurl <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'

if(!file.exists('repdata_data_StormData.csv.bz2')) {
  download.file(fileurl, 'repdata_data_StormData.csv.bz2')
  #unzip(zipfile='repdata_data_activity.zip')
  #file.remove('repdata_data_activity.zip')
  DATA <- read.csv("repdata_data_StormData.csv.bz2", header = TRUE, sep = ",")
} else {
  #unzip(zipfile='repdata_data_activity.zip')
  DATA <- read.csv("repdata_data_StormData.csv.bz2", header = TRUE, sep = ",")
}
```

The are `r nrow(DATA)` observations in this dataset, for the variables 
```{r , echo = TRUE, comment = ""}
head(DATA)
```

The type of events are reported are
```{r , echo = TRUE, comment = ""}
unique(DATA$EVTYPE)
```

# Data Processing: Re-define the type of events
Some events are reporte d with acronyms, typos, etc. We will cluster these a broader label. Our new general labels are "STORM", "FOG", "FLOOD", "COLD", "VOLCANO", "HOT" (weather), "FIRE". For instance, "STORM" would summarize "SHOWER","PRECIPITATION","HURRICANE","TYPHOON"... and basically any event related to raining.
```{r , echo = TRUE}
DATA$EVTYPE_modified<-DATA$EVTYPE
DATA$EVTYPE_modified<-trimws(as.character(DATA$EVTYPE_modified))
DATA$EVTYPE_modified<-toupper(as.character(DATA$EVTYPE_modified))

myValues <- c("SHOWER","PRECIPITATION","HURRICANE","TSTM","TSTMW","TUNDERSTORM","TYPHOON","WIND","WND","LIGHTNING","WINTER STORMS","STORM","GUSTS","HAIL","WHIRLWIND","CLOUD","TORNDAO","TORNADOES","TORNADO","RAIN", "PRECIPATATION", "PRECIP" )
pattern <- paste(myValues, collapse = "|")
DATA$EVTYPE_modified[grep(pattern,c(as.character(DATA$EVTYPE_modified)))]<-"STORM"

myValues <- c("SURF", "TSUNAMI" , "TIDE" )
pattern <- paste(myValues, collapse = "|")
DATA$EVTYPE_modified[grep(pattern,c(as.character(DATA$EVTYPE_modified)))]<-"TSUNAMI/TIDES"

myValues <- c("SLID" )
pattern <- paste(myValues, collapse = "|")
DATA$EVTYPE_modified[grep(pattern,c(as.character(DATA$EVTYPE_modified)))]<-"SLIDE*"

myValues <- c("FOG" )
pattern <- paste(myValues, collapse = "|")
DATA$EVTYPE_modified[grep(pattern,c(as.character(DATA$EVTYPE_modified)))]<-"FOG"

myValues <- c("FLOOD","FLOODING","FLD","WET","SPOUT","RAIN","WET","TIDAL")
pattern <- paste(myValues, collapse = "|")
DATA$EVTYPE_modified[grep(pattern,c(as.character(DATA$EVTYPE_modified)))]<-"FLOOD"

myValues <- c("FREEZING","HYPOTHERMIA","ICE","FREEZE","WINTRY","WINTERY","WINTER WEATHER","SNOW","COLD","LOW TEMP","COOL","COLD" ,"BLIZZARD")
pattern <- paste(myValues, collapse = "|")
DATA$EVTYPE_modified[grep(pattern,c(as.character(DATA$EVTYPE_modified)))]<-"COLD"

myValues <- c("VOLCANIC","VOG")
pattern <- paste(myValues, collapse = "|")
DATA$EVTYPE_modified[grep(pattern,c(as.character(DATA$EVTYPE_modified)))]<-"VOLCANO"

myValues <- c("HYPERTHERMIA","WARM","DRY","WARM","WARMTH","DRYHOT","HOT","HEAT")
pattern <- paste(myValues, collapse = "|")
DATA$EVTYPE_modified[grep(pattern,c(as.character(DATA$EVTYPE_modified)))]<-"HOT"

myValues <- c("WILDFIRES","FIRE","FIRES","WILDFIRE")
pattern <- paste(myValues, collapse = "|")
DATA$EVTYPE_modified[grep(pattern,c(as.character(DATA$EVTYPE_modified)))]<-"FIRE"

# Un-comment 
#unique(DATA$EVTYPE_modified)
#table(DATA$EVTYPE_modified)
```


## Exponent

PROPDMGEXP and CROPDMGEXP are used as exponents to the numeric values for the damage. 
The only symbols with a clear meaning are:
   
1. H or h:  hundreds of dollars
2. K or k:  thousands of dollars
3. M or m:  millions of dollars
4. B or b:  billions of dollars

We leave the symbols without a clear meaning without change. 

```{r, echo = TRUE, comment= ""}

table(DATA$PROPDMGEXP)

DATA$PROPDAMAGE = DATA$PROPDMG

DATA[DATA$PROPDMGEXP == "h", ]$PROPDAMAGE = DATA[DATA$PROPDMGEXP == "h", ]$PROPDMG * 10^2
DATA[DATA$PROPDMGEXP == "H", ]$PROPDAMAGE = DATA[DATA$PROPDMGEXP == "H", ]$PROPDMG * 10^2
DATA[DATA$PROPDMGEXP == "K", ]$PROPDAMAGE = DATA[DATA$PROPDMGEXP == "K", ]$PROPDMG * 10^3
DATA[DATA$PROPDMGEXP == "m", ]$PROPDAMAGE = DATA[DATA$PROPDMGEXP == "m", ]$PROPDMG * 10^6
DATA[DATA$PROPDMGEXP == "M", ]$PROPDAMAGE = DATA[DATA$PROPDMGEXP == "M", ]$PROPDMG * 10^6
DATA[DATA$PROPDMGEXP == "B", ]$PROPDAMAGE = DATA[DATA$PROPDMGEXP == "B", ]$PROPDMG * 10^9

table(DATA$CROPDMGEXP)

DATA$CROPDAMAGE = DATA$CROPDMG
DATA[DATA$CROPDMGEXP == "H", ]$CROPDAMAGE = DATA[DATA$CROPDMGEXP == "H", ]$CROPDMG * 10^2
DATA[DATA$CROPDMGEXP == "k", ]$CROPDAMAGE = DATA[DATA$CROPDMGEXP == "k", ]$CROPDMG * 10^3
DATA[DATA$CROPDMGEXP == "K", ]$CROPDAMAGE = DATA[DATA$CROPDMGEXP == "K", ]$CROPDMG * 10^3
DATA[DATA$CROPDMGEXP == "m", ]$CROPDAMAGE = DATA[DATA$CROPDMGEXP == "m", ]$CROPDMG * 10^6
DATA[DATA$CROPDMGEXP == "M", ]$CROPDAMAGE = DATA[DATA$CROPDMGEXP == "M", ]$CROPDMG * 10^6
DATA[DATA$CROPDMGEXP == "B", ]$CROPDAMAGE = DATA[DATA$CROPDMGEXP == "B", ]$CROPDMG * 10^9
```

# Results
We consider as "harmful to population health" the events with considerable number of fatalities or injuries. We choose 10 as a cutoff to define a harmful event to the public, as fewer might not represent a burden to public health centers.

```{r , echo = TRUE, fig.height= 6, fig.width= 10,fig.align = "center"}
#subset(x=DATA, DATA$FATALITIES>10 ||DATA$INJURIES>10 , select=c(FATALITIES,INJURIES, #EVTYPE_modified))
#table(subset(x=DATA, DATA$FATALITIES>10 |DATA$INJURIES>10 , select=c(FATALITIES,INJURIES, EVTYPE_modified)))

SUBDATA<- subset(x=DATA, DATA$FATALITIES>10|DATA$INJURIES>10 , select=c(FATALITIES,INJURIES, EVTYPE_modified))
total<-aggregate(SUBDATA[,1:2 ], list(SUBDATA$EVTYPE_modified), FUN=function(x){sum(x,na.rm=TRUE)})
colnames(total)<-c("EVTYPE","FATALITIES","INJURIES")

par(mar = c(10,10,4,2) + 1, par(las=2))
barplot(log(as.matrix(t(total[,2:3] +1)), base = 100), col=c("orange", "brown"),
        xlab= "log_100 number of incidents", axisnames = T ,xlim=c(0,3),
        names.arg  = as.vector(total$EVTYPE),beside = TRUE, horiz = T
)

legend(x=2,y=6, legend = colnames(total)[2:3], 
       fill = c("orange", "brown"), box.lty = 1, cex = 1)
```

To filter the dmage impact we would consider a damage with public impact if its cost are above 10 million dollars
```{r , echo = TRUE, fig.height= 6, fig.width= 10,fig.align = "center"}
SUBDATA<- subset(x=DATA, DATA$PROPDAMAGE>10^7|DATA$CROPDAMAGE>10^7 , select=c(PROPDAMAGE,CROPDAMAGE, EVTYPE_modified))
total<-aggregate(SUBDATA[,1:2 ], list(SUBDATA$EVTYPE_modified), FUN=function(x){sum(x,na.rm=TRUE)})
colnames(total)<-c("EVTYPE","PROPDAMAGE","CROPDAMAGE")

#pie(total$PROPDAMAGE, cex=1, labels = total$EVTYPE)
#pie(total$CROPDAMAGE, cex=1, labels = total$EVTYPE)
par(mar = c(10,10,4,2) + 0.1, par(las=2))
barplot(log(as.matrix(t(total[,2:3] +1)), base = 100), col=c("orange", "brown"),
        xlab= "log_100 number of incidents", axisnames = T , xlim=c(0,6),
        names.arg  = as.vector(total$EVTYPE),beside = TRUE, horiz = T
        )
legend(x=1,y=6, legend = colnames(total)[2:3], 
       fill = c("orange", "brown"), box.lty = 1, cex = 1)

```

# Conclusions
On one hand, we observe that the most harmful event for public health is `STORM` and `HOT` weather in terms of fatalities and injuries. On the other hand, `FLOOD` and `STORM` create the most property and crop damages.  
